---
title: "Personal expenses analysis"
date: "2026-03-26"
format: html
---

::: {.callout-note}
This report was generated using AI (Databot) under general human direction. At the time of generation, the contents have not been comprehensively reviewed by a human analyst.

<!--
To indicate human review: Delete the line above about contents not being reviewed, and replace this comment with:
The contents have been reviewed and validated by [Your Name], [Your Role] on [Date].
-->
:::

# Introduction

This report analyzes personal expense data loaded from `experiments/expenses/data/expenses.csv`. The dataset contains transactions across multiple accounts and categories, covering the period from March 2025 through January 2026. Non-expense entries (income, rental income, transfers, and investments) have been excluded so the analysis focuses on true spending.

```{r}
#| label: load-packages
library(tidyverse)
library(gt)
```

```{r}
#| label: load-data
expenses_raw <- read.csv("../../../expenses/data/expenses.csv")

# Drop unnecessary columns
expenses_raw <- expenses_raw[, !names(expenses_raw) %in% c("type", "id", "spend_type", "vendor")]

# Filter out non-expense categories
expenses <- expenses_raw |>
  filter(!category %in% c("income", "rental inc", "transfer/pmt", "investment"))
```

# Spending by category

The table below shows total spending per category, sorted from highest to lowest.

```{r}
#| label: spending-by-category
expenses |>
  group_by(category) |>
  summarise(total = sum(amount) * -1) |>
  arrange(desc(total))
```

The top three expense categories are food & bev, travel/transport, and rent, which together account for the majority of total spending.

# Monthly spending breakdown

The table below breaks down total spending by category and month, giving a view of how spending patterns shift over time.

```{r}
#| label: monthly-breakdown-table
monthly_wide <- expenses |>
  group_by(month, category) |>
  summarise(total = sum(amount) * -1, .groups = "drop") |>
  mutate(month = factor(month, levels = sort(unique(month)))) |>
  arrange(month) |>
  pivot_wider(names_from = month, values_from = total, values_fill = 0)

month_cols <- names(monthly_wide)[-1]

monthly_wide |>
  gt() |>
  fmt_currency(columns = all_of(month_cols), currency = "USD", decimals = 0) |>
  data_color(
    columns = all_of(month_cols),
    rows = everything(),
    method = "numeric",
    palette = c("white", "yellow", "red"),
    domain = c(0, max(select(monthly_wide, all_of(month_cols))))
  ) |>
  cols_label(category = "Category") |>
  tab_header(title = "Monthly spending by category") |>
  opt_stylize(style = 1)
```

Note that January 2026 (`01-2026`) appears to be a partial month and should be interpreted with caution.

# Monthly spending chart

The stacked bar chart below visualizes total spending by category for each month.

```{r}
#| label: monthly-spending-chart
expenses |>
  group_by(month, category) |>
  summarise(total = sum(amount) * -1, .groups = "drop") |>
  mutate(month = factor(month, levels = sort(unique(month)))) |>
  ggplot(aes(x = month, y = total, fill = category)) +
  geom_col() +
  scale_y_continuous(labels = scales::dollar) +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Monthly spending by category",
    x = "Month",
    y = "Total spending",
    fill = "Category"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

November and December 2025 stand out as the highest-spending months, likely driven by holiday and travel costs. May 2025 shows a notable spike in `int/bank chg/tax`, explained by a large tax payment (see below).

# Top 10 largest expenses

The table below lists the 10 largest individual transactions in the dataset.

```{r}
#| label: top-10-expenses
expenses |>
  arrange(amount) |>
  head(10) |>
  select(date, name, category, month, amount) |>
  mutate(amount = amount * -1)
```

Key observations:

- The largest single expense is a **$6,300.50 tax payment** (CANADA TXD) in May 2025, which explains the `int/bank chg/tax` spike that month.
- Two large **travel bookings** appear in late 2025: G Adventures ($5,566.40 in December) and Expedia ($3,319.20 in November), contributing to the high year-end spending.
- A **$2,239.38 Apple charge** in August 2025 is notably large for a business expense and may warrant further review.
- Monthly **rent payments** of $2,125.00 appear consistently throughout the year.

# Key takeaways

- **Food & bev** and **travel/transport** are the largest discretionary expense categories.
- **Rent** is a steady fixed cost of $2,125/month.
- A **CRA tax payment of $6,300.50** in May 2025 caused a one-time spike in the `int/bank chg/tax` category.
- **November and December 2025** were the highest-spending months overall, driven by travel bookings.
- The **Apple charge of $2,239.38** in August 2025 is an outlier worth investigating further.
- **January 2026** data is incomplete and should not be compared directly to other months.
